---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

###First load all necessary libraries
```{r}
library(data.table)
library(lubridate)
library(ggplot2)
library(RColorBrewer)
library(knitr)
```

###loading and preprocessing data
```{r}
setwd("C:/Users/Rafael/Documents/GitHub/RepData_PeerAssessment1")
unzip("activity.zip")
pa1 <- fread("activity.csv")
pa1[,date := ymd(date)]
```



## What is mean total number of steps taken per day?
### What is the distribution of total number of steps taken per day?
```{r}
stepsDay <- pa1[,.(steps = sum(steps)),by=date]
ggplot(data = stepsDay, aes(steps)) + 
    geom_histogram() +
    labs(title = "Steps taken per day distribution")
```

### What is the mean number of steps taken per weekday?
```{r}
pa1[,weekDay := wday(date, label = TRUE)]
meanSteps <- pa1[,.(steps = mean(steps, na.rm = TRUE)), by=weekDay][order(weekDay)]
ggplot(data = meanSteps, aes(weekDay,steps)) + 
    geom_col() +
    labs(title = "Mean Steps by weekday")
```

### What is the median number of steps taken per weekday?
```{r}

medianSteps <- pa1[,.(steps = as.numeric(median(steps, na.rm = TRUE))), by=weekDay][order(weekDay)]
ggplot(data = medianSteps, aes(weekDay,steps)) + 
    geom_col() +
    labs(title = "Median Steps by weekday")
noSteps <- pa1[steps == 0, .N]
allSteps <- pa1[,.N]
```


Median has value equal 0 for every weekday because the data set has more 0 than actual value when looking grouped by day.
Total 0 value on data set is `r noSteps`
Total number of row on data set is `r allSteps`


## What is the average daily activity pattern?

```{r}
avgStepsIntervals <- pa1[,.(steps = sum(steps, na.rm = TRUE)),by=interval]
ggplot(data = avgStepsIntervals, aes(interval,steps)) + 
    geom_line() +
    labs(title = "Average steps per intervals of 5 min")
maxSteps <- avgStepsIntervals[,max(steps)]
maxInterval <- avgStepsIntervals[steps == maxSteps, interval]
maxTime <- today() + minutes(maxInterval)
```

The 5-minute interval that has the most steps on average had `r maxSteps` steps in the interval `r maxInterval` that corresponds to `r hour(maxTime)`h `r minute(maxTime)`m of the day.





## Imputing missing values

### Calculating total number of missing values

```{r}
naSteps <- sum(!complete.cases(pa1))
```

1 - Total number of missing values `r naSteps`

### Filling missing values. Mean for that weekday in that 5 min interval.

```{r}
fullPa1 <- pa1[, index := .GRP, by = 1:nrow(pa1)]
fullPa1$steps <- sapply(fullPa1$index, function (x)
    ifelse(is.na(fullPa1$steps[fullPa1$index == x])
           , mean(fullPa1$steps[
               fullPa1$weekDay == fullPa1$weekDay[fullPa1$index == x] 
               & fullPa1$interval == fullPa1$interval[fullPa1$index == x]
               ], na.rm = TRUE)
           ,fullPa1$steps[fullPa1$index == x]
    )
)
```

### Histogram, mean and Median with filled missing values

```{r}
stepsDayFull <- fullPa1[,.(steps = sum(steps)),by=date]
ggplot(data = stepsDayFull, aes(steps)) + 
    geom_histogram() +
    labs(title = "Steps taken per day distribution - No missing data")
```


```{r}
meanStepsFull <- fullPa1[,.(steps = mean(steps, na.rm = TRUE)), by=weekDay][order(weekDay)]
ggplot(data = meanStepsFull, aes(weekDay,steps)) + 
    geom_col() +
    labs(title = "Mean Steps by weekday - No missing data")
```

```{r}
medianStepsFull <- fullPa1[,.(steps = median(steps, na.rm = TRUE)), by=weekDay][order(weekDay)]
ggplot(data = medianStepsFull, aes(weekDay,steps)) + 
    geom_col() +
    labs(title = "Median Steps by weekday - No missing data")
noStepsFull <- fullPa1[steps == 0, .N]
allStepsFull <- fullPa1[,.N]
```
Median has value equal 0 for every weekday because the data set has more 0 than actual value when looking grouped by day.
Total 0 value on data set is `r noStepsFull`
Total number of row on data set is `r allStepsFull`


### What is the impact of imputing missing data on the estimates of the total daily number of steps?

####Steps by day
```{r}
stepsComp <- stepsDay[stepsDayFull, on = "date"]
setnames(stepsComp, c("date", "steps with NA", "steps filled NA"))
stepsComp[,Difference := `steps filled NA`-ifelse(is.na(`steps with NA`),0,`steps with NA`)]
stepsCompDiff <- as.integer(round(stepsComp[,sum(Difference)],0))
```

The total difference from total steps taken from data with NA against NA filled is `r stepsCompDiff` steps. 



####Mean steps taken per weekday
```{r}
meanComp <- meanSteps[meanStepsFull, on = "weekDay"]
setnames(meanComp, c("weekDay", "steps with NA", "steps filled NA"))
meanComp[,Difference := `steps with NA`-`steps filled NA`]
meanCompDiff <- meanComp[,sum(Difference)]
```

The total difference from mean steps taken by weekday from data with NA against NA filled is `r meanCompDiff`.
This means that data has no difference since it was filled with the mean of same dimension.

####Median steps taken per weekday
```{r}
medianComp <- medianSteps[medianStepsFull, on = "weekDay"]
setnames(medianComp, c("weekDay", "steps with NA", "steps filled NA"))
medianComp[,Difference := `steps with NA`-`steps filled NA`]
medianCompDiff <- medianComp[,sum(Difference)]
```

The total difference from median steps taken by weekday from data with NA against NA filled is `r medianCompDiff`.
This means that data has no difference since it was filled with the median of same dimension.




## Are there differences in activity patterns between weekdays and weekends?

```{r}
fullPa1[,typeOfDay := ifelse(wday(date) > 6, "Weekend", "Weekday")]
fullPa1[,typeOfDay := as.factor(typeOfDay)]
table(fullPa1[,typeOfDay])
```

```{r}
meanFullPa1 <- fullPa1[, .(steps = mean(steps)), by=.(interval, typeOfDay)]
ggplot(data = meanFullPa1, aes(interval, steps))+
    geom_line()+
    facet_wrap(~typeOfDay, ncol=1)
```
